---
title: "scRNA-seq Tumor Microenvironment (Seurat)"
author: "Anastasia Ghilkovsky"
output: html_document
---

```{r}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 5)
```

Packages: 

# Install Packages 
install.packages("Seurat")
install.packages("tidyverse")
install.packages("patchwork")
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
BiocManager::install(c("org.Hs.eg.db", "clusterProfiler"))


```{r}
# Load libraries
library(Seurat)
library(tidyverse)
library(patchwork)
library(jsonlite)
library(tibble)
library(dplyr)
library(org.Hs.eg.db)
library(clusterProfiler)
library(reshape2)
library(ggplot2)
```


Importing Data for QC
```{r}

#Path to the raw_feature_bc_matrix folder
data_dir <- "scRNA_tumor/data/raw_feature_bc_matrix"


#Load data
tumor_data <- Read10X(data.dir = data_dir)

#Create Seurat object
seurat_obj <- CreateSeuratObject(counts = tumor_data, project = "Tumor_Mix", min.cells = 3, min.features = 200)


#Add metadata: percent mitochondrial genes
seurat_obj[["percent.mt"]] <- PercentageFeatureSet(seurat_obj, pattern = "^MT-")

```

Violin Plots for QC
```{r}
VlnPlot(seurat_obj, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)


seurat_obj <- subset(seurat_obj, 
                     subset = nFeature_RNA > 500 & nFeature_RNA < 6000 & percent.mt < 15)

seurat_obj

seurat_obj <- NormalizeData(seurat_obj, normalization.method = "LogNormalize", scale.factor = 10000)

seurat_obj <- FindVariableFeatures(seurat_obj, selection.method = "vst", nfeatures = 2000)


top10 <- head(VariableFeatures(seurat_obj), 10)
```


Create Base Plot
```{r}
plot1 <- VariableFeaturePlot(seurat_obj)

#Add top 10 labels to plot
plot1 + LabelPoints(plot = plot1, points = top10, repel = TRUE)


seurat_obj <- ScaleData(seurat_obj, features = VariableFeatures(seurat_obj))

seurat_obj <- RunPCA(seurat_obj, features = VariableFeatures(seurat_obj))


print(seurat_obj[["pca"]], dims = 1:2, nfeatures = 10)

```

Elbow Plot
```{r}
ElbowPlot(seurat_obj)
```

Clustering
```{r}
seurat_obj <- FindNeighbors(seurat_obj, dims = 1:15)
seurat_obj <- FindClusters(seurat_obj, resolution = 0.5)  # try 0.4–1.2 for resolution tuning


seurat_obj <- RunUMAP(seurat_obj, dims = 1:15)

#Plot UMAP with cluster labels
DimPlot(seurat_obj, reduction = "umap", label = TRUE, pt.size = 0.5)


cluster_markers <- FindAllMarkers(seurat_obj, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)

#Top markers per cluster
top_markers <- cluster_markers %>% group_by(cluster) %>% top_n(n = 5, wt = avg_log2FC)

#View results
head(top_markers)
```

Feature Plot
```{r}
FeaturePlot(seurat_obj, features = c("CD3D", "CD14", "EPCAM", "MS4A1", "LYZ", "MKI67"))
```

Top Markers, Cell-Type Annotation
```{r}
top_markers <- cluster_markers %>% group_by(cluster) %>% top_n(n = 3, wt = avg_log2FC)
top_markers

new_cluster_ids <- c(
  "Macrophages", "Tumor cells", "Tumor cells", "B cells", "T cells", "Monocytes", 
  "T cells", "B cells", "Tumor cells", "NK cells", "T cells", "DCs", 
  "T cells", "T cells", "Tumor cells", "Tumor cells", "DCs", "T cells", 
  "Prolif. cells", "B cells", "Stromal", "Unknown"
)

#Assign new identities
names(new_cluster_ids) <- levels(seurat_obj)
seurat_obj <- RenameIdents(seurat_obj, new_cluster_ids)

DimPlot(seurat_obj, reduction = "umap", label = TRUE, pt.size = 0.5) + NoLegend()

head(seurat_obj@meta.data)
```

Sample Type Assignment (JSON)
```{r}
#Load libraries
library(jsonlite)
library(tibble)
library(dplyr)
library(Seurat)

#Set the path to JSON file
tag_path <- "scRNA_tumor/data/cells_per_tag.json"


#Read JSON
tag_data <- fromJSON(tag_path)

#Convert JSON list to long format tibble
tag_df <- bind_rows(lapply(names(tag_data), function(tag) {
  tibble(barcode = unlist(tag_data[[tag]]), sample = tag)
}))

#Match barcodes to cell names in Seurat object
seurat_obj$SampleTag <- tag_df$sample[match(Cells(seurat_obj), tag_df$barcode)]

#Confirm assignment
cat("Number of cells with assigned sample tags:\n")
print(sum(!is.na(seurat_obj$SampleTag)))

#View distribution of sample tags
cat("Distribution of SampleTag assignments:\n")
print(table(seurat_obj$SampleTag, useNA = "always"))

saveRDS(seurat_obj, file = "seurat_with_sample_tags.rds")


DimPlot(seurat_obj, group.by = "SampleTag", reduction = "umap", label = TRUE) + 
  ggtitle("UMAP by Sample Tag")


Idents(seurat_obj) <- "SampleTag"
sample_markers <- FindAllMarkers(seurat_obj, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)
head(sample_markers)
```

Plots and Graphs 
```{r}
#Create Violin Plot
VlnPlot(seurat_obj, features = "CD3D", group.by = "SampleTag")


seurat_obj$SampleTag <- ifelse(is.na(seurat_obj$SampleTag), "Unassigned", seurat_obj$SampleTag)

DimPlot(seurat_obj, group.by = "SampleTag", reduction = "umap", label = TRUE) + 
  ggtitle("UMAP Colored by Sample")


top_genes <- sample_markers %>%
  group_by(cluster) %>%
  top_n(n = 5, wt = avg_log2FC)


#Create Heatmap
DoHeatmap(seurat_obj, features = top_genes$gene, group.by = "SampleTag") + 
  ggtitle("Top Markers by Sample")

DoHeatmap(
  seurat_obj,
  features = unique(top_genes$gene),
  group.by = "SampleTag",
  raster = FALSE  # higher resolution for export
) +
  theme(
    axis.text.y = element_text(size = 10),
    plot.title = element_text(hjust = 0.5, face = "bold")
  ) +
  ggtitle("Top Sample-Specific Markers")


Idents(seurat_obj) <- "SampleTag"
de_genes <- FindMarkers(seurat_obj, ident.1 = "BC002", ident.2 = "BC001", logfc.threshold = 0.25)
head(de_genes)


VlnPlot(seurat_obj, features = c("SLPI", "PAX8", "IGFBP3"), group.by = "SampleTag", pt.size = 0)

FeaturePlot(seurat_obj, features = c("SLPI", "PAX8"), split.by = "SampleTag")

write.csv(de_genes, file = "DEG_BC002_vs_BC001.csv")

top_up <- head(de_genes[order(-de_genes$avg_log2FC), ], 10)
top_down <- head(de_genes[order(de_genes$avg_log2FC), ], 10)
genes_to_plot <- rownames(rbind(top_up, top_down))

DoHeatmap(
  seurat_obj,
  features = genes_to_plot,
  group.by = "SampleTag",
  size = 3
) +
  scale_fill_gradientn(colors = c("magenta", "black", "yellow"), limits = c(0, 2.5)) +
  theme(
    axis.text.x = element_text(size = 6, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 6),
    axis.title = element_blank(),
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA),
    legend.background = element_rect(fill = "white", color = NA),
    legend.key = element_rect(fill = "white", color = NA),
    panel.grid = element_blank()
  )


all_markers <- FindAllMarkers(seurat_obj, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25)
head(all_markers)
write.csv(all_markers, "all_sample_markers.csv", row.names = FALSE)

top_genes <- all_markers %>% group_by(cluster) %>% top_n(n = 3, wt = avg_log2FC)
DotPlot(seurat_obj, features = unique(top_genes$gene), group.by = "SampleTag") + RotatedAxis()

top_deg <- de_genes %>% 
  filter(p_val_adj < 0.05 & abs(avg_log2FC) > 1) %>%
  rownames_to_column("gene") %>%
  pull(gene)
saveRDS(seurat_obj, file = "seurat_obj_with_sampletag.rds")


#Make sure the identities are set correctly
Idents(seurat_obj) <- "SampleTag"

#FindAllMarkers to get markers per sample
all_markers <- FindAllMarkers(
  seurat_obj,
  only.pos = TRUE,          # only return positive markers
  min.pct = 0.25,           # expressed in at least 25% of cells
  logfc.threshold = 0.25    # minimum log fold change
)

#View the top rows
head(all_markers)

write.csv(all_markers, "all_sample_markers.csv", row.names = FALSE)

#Load libraries
library(dplyr)

top_markers <- all_markers %>%
  group_by(cluster) %>%
  top_n(n = 3, wt = avg_log2FC)

top_genes <- unique(top_markers$gene)

DotPlot(seurat_obj, features = top_genes, group.by = "SampleTag") + RotatedAxis()


DimPlot(seurat_obj, group.by = "SampleTag", reduction = "umap", label = TRUE, pt.size = 0.5)

FeaturePlot(seurat_obj, features = c("SLPI", "PAX8", "IGFBP3"), cols = c("lightgrey", "red"), reduction = "umap")


#Save Seurat object
saveRDS(seurat_obj, file = "seurat_with_sample_tags.rds")

#Save current heatmap
ggsave("top_markers_heatmap.png", width = 10, height = 8, dpi = 300)

immune_markers <- list(
  T_cells = c("CD3D", "CD3E"),
  B_cells = c("MS4A1"),
  NK_cells = c("NKG7", "GNLY"),
  Macrophages = c("CD68", "CD163"),
  Dendritic_cells = c("FCER1A", "CST3")
)

immune_scores <- AddModuleScore(
  object = seurat_obj,
  features = immune_markers,
  name = names(immune_markers)
)

library(ggplot2)
library(reshape2)

score_df <- as.data.frame(seurat_obj@meta.data[, grep("_1", colnames(seurat_obj@meta.data))])
score_df$SampleTag <- seurat_obj$SampleTag

melted <- melt(score_df, id.vars = "SampleTag")
ggplot(melted, aes(x = SampleTag, y = value, fill = variable)) +
  geom_bar(stat = "summary", fun = "mean", position = "dodge") +
  labs(title = "Immune cell type scores per sample", y = "Average module score", x = "Tumor Sample") +
  theme_minimal()

head(melted)
str(melted)

#Rename columns to their actual immune cell names
score_columns <- grep("_1$", colnames(seurat_obj@meta.data), value = TRUE)
names(score_columns) <- names(immune_markers) 
colnames(seurat_obj@meta.data)[colnames(seurat_obj@meta.data) %in% score_columns] <- names(score_columns)

head(colnames(seurat_obj@meta.data), 20)

#Add module scores for each immune cell type
for (cell_type in names(immune_markers)) {
  seurat_obj <- AddModuleScore(
    object = seurat_obj,
    features = list(immune_markers[[cell_type]]),
    name = cell_type
  )
}


#Get new added module score columns
score_columns <- grep("1$", colnames(seurat_obj@meta.data), value = TRUE)

#Add SampleTag for grouping
score_df <- seurat_obj@meta.data[, c(score_columns, "SampleTag")]

#Melt the data frame for ggplot
library(reshape2)
melted <- melt(score_df, id.vars = "SampleTag")


library(ggplot2)

ggplot(melted, aes(x = SampleTag, y = value, fill = variable)) +
  geom_bar(stat = "summary", fun = "mean", position = "dodge") +
  labs(title = "Immune Cell Type Scores per Tumor Sample",
       y = "Average Module Score", x = "Tumor Sample") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


#Check high level tumor lineage markers
FeaturePlot(seurat_obj, features = c("EPCAM", "KRT8", "PAX8", "SLPI", "IGFBP3", "MUC16", "CA125", "CDH1"))

#ViolinPlot across SampleTag
VlnPlot(seurat_obj, features = c("PAX8", "SLPI", "IGFBP3"), group.by = "SampleTag")
```

GO and KEGG
```{r}
Idents(seurat_obj) <- "SampleTag"
bc001_vs_others <- FindMarkers(seurat_obj, ident.1 = "BC001", ident.2 = c("BC002", "BC003", "BC004"), min.pct = 0.25, logfc.threshold = 0.25)

up_genes <- rownames(bc001_vs_others[bc001_vs_others$avg_log2FC > 0.5 & bc001_vs_others$p_val_adj < 0.05, ])

if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
BiocManager::install(c("org.Hs.eg.db", "clusterProfiler"))



library(org.Hs.eg.db)
library(clusterProfiler)

Idents(seurat_obj) <- "seurat_clusters"

cluster_markers <- FindAllMarkers(seurat_obj, 
                                  only.pos = TRUE, 
                                  min.pct = 0.25, 
                                  logfc.threshold = 0.25)


cluster4_genes <- cluster_markers %>% filter(cluster == "4") %>% pull(gene)


library(org.Hs.eg.db)
library(clusterProfiler)

cluster4_entrez <- bitr(cluster4_genes, 
                        fromType = "SYMBOL", 
                        toType = "ENTREZID", 
                        OrgDb = org.Hs.eg.db)


go_results <- enrichGO(gene         = cluster4_entrez$ENTREZID,
                       OrgDb        = org.Hs.eg.db,
                       keyType      = "ENTREZID",
                       ont          = "BP",
                       pAdjustMethod = "BH",
                       pvalueCutoff = 0.05,
                       qvalueCutoff = 0.05)


kegg_results <- enrichKEGG(gene         = cluster4_entrez$ENTREZID,
                           organism     = 'hsa',
                           pvalueCutoff = 0.05)



barplot(go_results, showCategory = 10, title = "GO Enrichment")
dotplot(kegg_results, showCategory = 10, title = "KEGG Pathways")
```


